Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-24 analysis/paper/paper.Rmd
Added a CA excluding the non-flint material

This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as procurement strategies and access to raw materials. Finally, analytic and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification within larger unified projects.

@@

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon probability density functions are given as the posterior sum. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 10, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

(overview + sitemap) /
(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@bjerck2010; @glørstad2010]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good chronological control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

The data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. An in depth-analysis of the flint quality from sevens sites did result in the finding that the flint quality was relatively high, and consistently so across the sites. However, despite flint being treated as a unified category here, the cited study by was based on only seven sites located some km distance from the present study area, meaning that the results might not necessarily be transferable to the sites treated here [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @glørstad2011; @eigeland2015, 370].

Studies concerned with chronological changes in the composition of lithic assemblages in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @damlien2016; @berg-hansen2017; @solheim2020]. Studies that have involved the entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018; @breivik2020], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016]. These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledege, only a single study in southern Norway, concerned with the analysis of eight Middle Mesolithic assemblages, has involved the use of a multivariate quantitative framework to structure the analysis [@solheim2013a]. In sum, therefore, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult to consistently balance the comparison of a larger number of assemblages.

@@

The relatively constrained geographical distribution of the analysed sites, the limited temporal spread over which they where investigated, as well as the methodological equivalency across projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified. However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported on here, the data and employed R programming script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017]. However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves what Binford termed a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon. The process can still provide the identification of empirical patterns along the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses and lay the foundation for a deductive research agenda for which clear test implications can be derived [@clark2009, 29].

The first part of the analysis involved employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise was to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken was also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While largely intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, for the most part we lack even a most basic understanding of what any individual lithic object in an assemblage has been used for [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].

However, following not least from the ambiguous definition first put forward by @binford1979, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g. @bamforth1986; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1999 and collegues [e.g. @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation. This is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the sites [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].

These are the main measure employed here. However, as these measures are argued to predominantly be determined by mobility patterns, some additional variables suggested by @Bicho2020 to be related to mobility patterns are also explored. This is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf. @baxter1994, 100].

A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here. This pertains to number of site features such as fireplaces, cooking pits and dwelling structures. Feature count has been omitted as taphonomic loss is likely to have led to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied. This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].

@@

```{r lithics}
lithics <- sites %>%
  # Combine flint projectiles into a single variable
  mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
    macro_tool = axe + chisel, 
    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
    small_tool_other = knife_other + scraper_other + drill_other + 
      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
      flake_retouch_other,
    .keep = "unused") %>%
  # Make site names the rownames
  column_to_rownames(var = "n") %>% 
  # Deselect a few of the columns not be used
  select(-excavation_project, -excavated_m3, -dating_method)

@@

```{r co_data}
# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
# so is excluded here
calithics <- CA(sqrt(select(lithics, -site, -mean_date, -artefact_count, -flint_count, 
                       -nonflint_count, -reported_start_bc, -reported_end_bc,
                       -askeladden_id)),
                graph = FALSE)

@@

Figure \@ref(fig:cor) displays the CA using the lithic count data. The square root has been taken of the data before running the analysis to reduce the impact of extreme values that were making the main cluster of sites in the bottom right of \@ref(fig:cor)A difficult to discern [cf. @lockyear2000]. The general impression from the plots is that a chronological dimension is associated with the patterning in the data. This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data. The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension. Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to larger extent are characterised by core fragments, both in flint and non-flint materials, than cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. The single yellow point in the upper right quadrant of the row plot is Nedre Hobekk 2, which represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite. The use of metarhyolite and smoky quartz for the production of axes is present at other sites as well, but is evidently not as prominent a part of these assemblages. In sum, the findings for the earliest sites are in large part in line with previous research [e.g. @bang-andersen2003], although perhaps highlighting the relative importance of non-flint materials in south-eastern which, while also present, is typically not as marked in other parts of southern Norway [@breivik2018, 184]

>>>>

This study employs multivariate exploratory statistics to analyse lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as procurement strategies and access to raw materials. Finally, analytic and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification within larger unified projects.

@@

```{r 1dist, fig.height = 10, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

(overview + sitemap) 
# (temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))

@@

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon age determintations are given as the sum of the posterior density esimates. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 9, fig.width = 12}
temp_dist
```

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@bjerck2010]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g. @åstveit2018; @breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good chronological control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year towards the end of the Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

The dates associated with the sites are given. Where radiocarbon age determinations are available these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam]. The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function. The resulting posterior density estimate was then summed for each site.

The lithic data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together here have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @glørstad2011; @eigeland2015, 370]. Finally, while factors such as landscape changes through shoreline displacement still can have led to variable raw-material availability on the analysed sites, the relatively constrained geographical distribution of the sites hopefully counteracts some non-behavioural sources of variation.

Studies concerned with chronological changes in the composition of lithic assemblages in Norway have typically had a focus on formal tool types [e.g. @helskog1976; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @damlien2016; @berg-hansen2017; @solheim2020]. Studies that have involved the entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018; @breivik2020], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016]. These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledege, only a single study dealing with Mesolithic material in southern Norway has involved the use of a multivariate quantitative framework to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that are difficult to up-scale and consistently balance in the comparison of a larger number of assemblages.

@@

The relatively constrained geographical distribution of the analysed sites, the limited temporal spread over which they where investigated, as well as the methodological equivalency across projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified. However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported on here, the data and employed R programming script is freely available as a research compendium following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017]. However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves what Binford termed a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon. The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses. This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies --- studies that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken was also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapsed for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While largely intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, for the most part we lack even a most basic understanding of what any individual lithic object in an assemblage has been used for [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].

However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g. @bamforth1986; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1999 and collegues [e.g. @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation. This is in turn mainly to follow from the accumulated nature of land-use, procurement and mobility patterns associated with the sites [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017]. These are the main measure employed here. However, as these measures are argued to predominantly be determined by mobility patterns, the investigation of @Bicho2020 has also led to the inclusion of relative frequency of chips in the analysis as well, as they argue that this is an important. The analysed variables are therefore relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched and ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per m³; as well as relative frequency of chips, defined as artefacts with size \< 0.1 cm; relative frequency of cores which are simply all artefacts classified as cores in the original reports, as well as relative frequency blanks, here constituted by the categories flakes, blades, micro-blades and fragments. The analysis is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf. @baxter1994, 100].

A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here. This pertains to number of site features such as fireplaces, cooking pits and dwelling structures. Feature count has been omitted as taphonomic loss is likely to have led to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied. This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].

@@

```{r lithics}
lithics <- sites %>%
  # Combine flint projectiles into a single variable
  mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
    macro_tool = axe + chisel, 
    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
    small_tool_other = knife_other + scraper_other + drill_other + 
      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
      flake_retouch_other,
    .keep = "unused") %>%
  # Make site names the rownames
  column_to_rownames(var = "no") %>% 
  # Deselect a few of the columns not be used
  select(-excavation_project, -excavated_m3, -dating_method)

@@

```{r co_data}
# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
# so is excluded here
calithics <- CA(select(lithics, -site, -mean_date, -artefact_count, -flint_count, 
                       -nonflint_count, -reported_start_bc, -reported_end_bc,
                       -askeladden_id),
                graph = FALSE)

caflint <- CA(select(flint, -site, -mean_date, -artefact_count, -flint_count, 
                       -nonflint_count, -reported_start_bc, -reported_end_bc,
                       -askeladden_id),
                graph = FALSE)

@@

Figure \@ref(fig:cor) displays the CA using the lithic count data. The general impression from the plots is that a chronological dimension is associated with the patterning in the data. This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the clear horse-shoe shaped structure in the column plot [e.g. @lockyear2000], combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data. The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension. Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014] . The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages. In sum, the findings for the earliest sites are in large part in line with previous research [e.g. @bang-andersen2003; @damlien2018b], although perhaps highlighting the relative importance of non-flint materials for some of the sites in south-eastern Norway which, while also present, is typically not as marked in other parts of southern Norway [@breivik2018, 184].

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages. Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370, @glørstad2011].
